---
title: "Training and Comparing ML Classifiers on IMDb Reviews"
author: "Felix Lennert"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    theme: cosmo
freeze: true
jupyter: bert-env
---

## Description and motivation

I taught a graduate Computational Social Science course at Leipzig University. We focused heavily on quantitative text analysis. At some point, inevitably the question came up: what are the tradeoffs in terms of speed and quality. Thus, in the following document I compare supervised text classification approaches: dictionary-based, bag-of-words models, fine-tuned BERT, and zero-shot prompting with a local LLM. Using 500 training examples from IMDb reviews, I show how preprocessing and model choice affect accuracy, precision, recall, F1, and runtime.

## Data source

IMDb movie reviews with binary sentiment labels (positive/negative) from [Kaggle](https://www.kaggle.com/datasets/lakshmi25npathi/imdb-dataset-of-50k-movie-reviews). Data are not published; the code below assumes the CSV lives in `files_for_project_backlog/ml_classifiers/imdb-reviews.csv`.

## Outline

We load the data, split 500 train / remainder test (stratified), and run: (1) dictionary (AFINN), (2) bag-of-words (Naive Bayes, logistic, Lasso, Ridge, SVM, random forest, XGBoost), (3) BERT, (4) LLM zero-shot. Preprocessing is tokenization, stop-word removal, token filter (top 1,000), TF-IDF; SVM adds normalization.

## Analysis

### Setup and data split

```{r}
#| label: setup
#| output: false

needs(tidyverse, tidytext, textrecipes, tidymodels, workflows, yardstick, discrim,
      naivebayes, SnowballC, stopwords, textdata, tictoc, jsonlite,
      glmnet, LiblineaR, ranger, xgboost, here)

base <- here("files_for_project_backlog", "ml_classifiers")
imdb_raw <- read_csv(here(base, "imdb-reviews.csv"), show_col_types = FALSE) |>
  rename(text = review)

set.seed(123)
split <- initial_split(imdb_raw, prop = 500 / nrow(imdb_raw), strata = sentiment)
imdb_train <- training(split)
imdb_test <- testing(split)

imdb_train |> write_csv(here(base, "train_split.csv"))
imdb_test |> write_csv(here(base, "test_split.csv"))
```

### Recipes

```{r}
#| label: recipes
#| output: false

recipe_bow <- recipe(sentiment ~ text, data = imdb_train) |>
  step_tokenize(text) |>
  step_tokenfilter(text, max_tokens = 1000) |>
  step_tfidf(text)

recipe_svm <- recipe(sentiment ~ text, data = imdb_train) |>
  step_tokenize(text) |>
  step_tokenfilter(text, max_tokens = 1000) |>
  step_tfidf(text) |>
  step_normalize(all_predictors())

run_bow <- function(name, wf, label_tic = name) {
  tic(label_tic)
  fit_wf <- fit(wf, data = imdb_train)
  pred <- augment(fit_wf, imdb_test)
  elapsed <- toc(quiet = TRUE)
  pred |>
    metric_set(accuracy, precision, recall, f_meas)(
      truth = sentiment,
      estimate = .pred_class,
      event_level = "second"
    ) |>
    pivot_wider(names_from = .metric, values_from = .estimate) |>
    mutate(classifier = name, runtime_seconds = round(elapsed, 2), .before = 1) |>
    rename(f1 = f_meas) |>
    select(classifier, accuracy, precision, recall, f1, runtime_seconds)
}
```

### Dictionary (AFINN)

Tokenize, stem, match AFINN lexicon, average score per review; classify by sign.

```{r}
#| label: dict

tic("Dictionary (AFINN)")
afinn_stemmed <- get_sentiments("afinn") |>
  mutate(stemmed = wordStem(word, language = "en"))
scored <- imdb_test |>
  rowid_to_column("doc") |>
  unnest_tokens(token, text) |>
  anti_join(get_stopwords(), by = c("token" = "word")) |>
  mutate(stemmed = wordStem(token, language = "en")) |>
  inner_join(afinn_stemmed, by = "stemmed", relationship = "many-to-many") |>
  group_by(doc) |>
  summarise(sentiment_score = mean(value), .groups = "drop") |>
  mutate(
    .pred_class = if_else(sentiment_score > 0, "positive", "negative"),
    .pred_class = factor(.pred_class, levels = c("positive", "negative"))
  )
imdb_afinn <- imdb_test |>
  rowid_to_column("doc") |>
  select(doc, sentiment) |>
  mutate(sentiment = factor(sentiment, levels = c("positive", "negative"))) |>
  left_join(select(scored, doc, .pred_class), by = "doc") |>
  mutate(
    .pred_class = replace_na(as.character(.pred_class), "positive"),
    .pred_class = factor(.pred_class, levels = c("positive", "negative"))
  )
dict_time <- toc(quiet = TRUE)
dict_metrics <- imdb_afinn |>
  metric_set(accuracy, precision, recall, f_meas)(
    truth = sentiment,
    estimate = .pred_class,
    event_level = "second"
  ) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  mutate(classifier = "Dictionary (AFINN)", runtime_seconds = round(dict_time, 2), .before = 1) |>
  rename(f1 = f_meas) |>
  select(classifier, accuracy, precision, recall, f1, runtime_seconds)
```

### Bag-of-words models

Same TF-IDF preprocessing; SVM adds normalization.

```{r}
#| label: bow

wf_nb <- workflow() |> add_recipe(recipe_bow) |> add_model(
  naive_Bayes() |> set_engine("naivebayes") |> set_mode("classification")
)
wf_glm <- workflow() |> add_recipe(recipe_bow) |> add_model(
  logistic_reg() |> set_engine("glm") |> set_mode("classification")
)
wf_lasso <- workflow() |> add_recipe(recipe_bow) |> add_model(
  logistic_reg(penalty = 0.01, mixture = 1) |> set_engine("glmnet") |> set_mode("classification")
)
wf_ridge <- workflow() |> add_recipe(recipe_bow) |> add_model(
  logistic_reg(penalty = 0.01, mixture = 0) |> set_engine("glmnet") |> set_mode("classification")
)
wf_svm <- workflow() |> add_recipe(recipe_svm) |> add_model(
  svm_linear() |> set_engine("LiblineaR") |> set_mode("classification")
)
wf_rf <- workflow() |> add_recipe(recipe_bow) |> add_model(
  rand_forest(trees = 100) |> set_engine("ranger") |> set_mode("classification")
)
wf_xgb <- workflow() |> add_recipe(recipe_bow) |> add_model(
  boost_tree(trees = 50) |> set_engine("xgboost") |> set_mode("classification")
)

bow_metrics <- bind_rows(
  run_bow("Naive Bayes (TF-IDF)", wf_nb),
  run_bow("Logistic regression (glm)", wf_glm),
  run_bow("Lasso (glmnet)", wf_lasso),
  run_bow("Ridge (glmnet)", wf_ridge),
  run_bow("SVM linear (LiblineaR)", wf_svm),
  run_bow("Random forest (ranger)", wf_rf),
  run_bow("XGBoost", wf_xgb)
)
```

### BERT (Python)

Fine-tuned `bert-base-uncased` on 500 samples. Writes metrics to `bert_results.json` for the R chunk below.

```{python}
#| label: bert
#| output: false

import json
import os
import time
from pathlib import Path

import pandas as pd
import torch
from torch import nn
from torch.utils.data import Dataset, DataLoader
from transformers import BertTokenizer, BertModel
from sklearn.metrics import accuracy_score, precision_score, recall_score, f1_score

# Find base: Quarto runs from project root; cwd may vary
for p in [Path.cwd(), Path.cwd().parent]:
  candidate = p / "files_for_project_backlog" / "ml_classifiers"
  if (candidate / "train_split.csv").exists():
    base = candidate
    break
else:
  base = Path.cwd() / "files_for_project_backlog" / "ml_classifiers"

train_df = pd.read_csv(base / "train_split.csv")
test_df = pd.read_csv(base / "test_split.csv")
if "review" in train_df.columns:
  train_df = train_df.rename(columns={"review": "text"})
if "review" in test_df.columns:
  test_df = test_df.rename(columns={"review": "text"})

label_map = {"negative": 0, "positive": 1}
X_train = train_df["text"].tolist()
y_train = [label_map[str(s).strip().lower()] for s in train_df["sentiment"]]
X_test = test_df["text"].tolist()
y_test = [label_map[str(s).strip().lower()] for s in test_df["sentiment"]]

class SentenceDataset(Dataset):
  def __init__(self, texts, labels, tokenizer, max_len=128):
    self.texts, self.labels, self.tokenizer, self.max_len = texts, labels, tokenizer, max_len
  def __len__(self):
    return len(self.texts)
  def __getitem__(self, idx):
    enc = self.tokenizer.encode_plus(str(self.texts[idx]), max_length=self.max_len,
      padding="max_length", truncation=True, return_tensors="pt")
    return {"input_ids": enc["input_ids"].flatten(), "attention_mask": enc["attention_mask"].flatten(),
            "label": torch.tensor(self.labels[idx], dtype=torch.long)}

class BertClassifier(nn.Module):
  def __init__(self, dropout=0.1):
    super().__init__()
    self.bert = BertModel.from_pretrained("bert-base-uncased")
    self.dropout = nn.Dropout(dropout)
    self.classifier = nn.Linear(768, 2)
  def forward(self, input_ids, attention_mask):
    out = self.bert(input_ids=input_ids, attention_mask=attention_mask)[1]
    return self.classifier(self.dropout(out))

device = torch.device("mps" if torch.backends.mps.is_available() else "cuda" if torch.cuda.is_available() else "cpu")
tokenizer = BertTokenizer.from_pretrained("bert-base-uncased")
train_ds = SentenceDataset(X_train, y_train, tokenizer)
test_ds = SentenceDataset(X_test, y_test, tokenizer)
train_loader = DataLoader(train_ds, batch_size=16, shuffle=True)
test_loader = DataLoader(test_ds, batch_size=16)

model = BertClassifier().to(device)
opt = torch.optim.AdamW(model.parameters(), lr=2e-5)
criterion = nn.CrossEntropyLoss()
t0 = time.perf_counter()
for _ in range(3):
  model.train()
  for batch in train_loader:
    opt.zero_grad()
    out = model(batch["input_ids"].to(device), batch["attention_mask"].to(device))
    criterion(out, batch["label"].to(device)).backward()
    opt.step()
model.eval()
preds, labels = [], []
with torch.no_grad():
  for batch in test_loader:
    out = model(batch["input_ids"].to(device), batch["attention_mask"].to(device))
    preds.extend(out.argmax(1).cpu().tolist())
    labels.extend(batch["label"].tolist())
elapsed = time.perf_counter() - t0

bert_results = {
  "accuracy": float(accuracy_score(labels, preds)),
  "precision": float(precision_score(labels, preds, average="weighted", zero_division=0)),
  "recall": float(recall_score(labels, preds, average="weighted", zero_division=0)),
  "f1": float(f1_score(labels, preds, average="weighted", zero_division=0)),
  "runtime_seconds": round(elapsed, 2)
}
(base / "bert_results.json").write_text(json.dumps(bert_results, indent=2))
```

### BERT metrics and LLM zero-shot

```{r}
#| label: bert-llm

bert_metrics <- jsonlite::read_json(here(base, "bert_results.json")) |>
  as.list() |>
  as_tibble() |>
  mutate(classifier = "BERT (fine-tuned)", .before = 1)

needs(ellmer, ollamar)
n_gpt <- 100
imdb_test_gpt <- imdb_test |> slice_head(n = n_gpt)
sentiment_type <- ellmer::type_object(
  sentiment = ellmer::type_enum("Sentiment", values = c("positive", "negative"))
)
prompt <- "You are a movie expert. Classify the review as positive or negative. Reply with one word: positive or negative."

tic("LLM zero-shot")
gpt_preds <- map(imdb_test_gpt$text, function(txt) {
  ch <- ellmer::chat_ollama(model = "qwen2.5:3b", system_prompt = prompt,
    params = ellmer::params(temperature = 0.1, seed = 42))
  as.character(ch$chat_structured(txt, type = sentiment_type)$sentiment)
}, .progress = TRUE)
gpt_time <- toc(quiet = TRUE)

gpt_df <- imdb_test_gpt |>
  select(sentiment) |>
  mutate(
    .pred_class = factor(unlist(gpt_preds), levels = c("positive", "negative")),
    sentiment = factor(sentiment, levels = c("positive", "negative"))
  )
gpt_metrics <- gpt_df |>
  metric_set(accuracy, precision, recall, f_meas)(
    truth = sentiment,
    estimate = .pred_class,
    event_level = "second"
  ) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  mutate(
    classifier = paste0("LLM zero-shot (n=", n_gpt, ")"),
    runtime_seconds = round(gpt_time, 2),
    .before = 1
  ) |>
  rename(f1 = f_meas) |>
  select(classifier, accuracy, precision, recall, f1, runtime_seconds)
```

## Results

```{r}
#| label: results
#| tbl-cap: "Accuracy, precision, recall, F1, and runtime per classifier. Train size 500 (LLM zero-shot)."

results <- bind_rows(dict_metrics, bow_metrics, bert_metrics, gpt_metrics)
results
```

## Conclusions

Stemming and token filtering improve dictionary and bag-of-words baselines. With 500 examples, TF-IDF + penalized regression or tree models outperform the dictionary; BERT reaches highest accuracy at higher compute cost. The zero-shot LLM (Qwen 2.5 3B) performs well without training and is flexible—change the prompt instead of retraining. Use dictionary or Naive Bayes for quick baselines; BERT when you have labels and GPU; LLM prompting when labels are scarce or categories change often.
